---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# pca

<!-- badges: start -->
[![R-CMD-check](https://github.com/asmauger/bios625hw3/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/asmauger/bios625hw3/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/asmauger/bios625hw3/branch/main/graph/badge.svg)](https://app.codecov.io/gh/asmauger/bios625hw3?branch=main)
<!-- badges: end -->

The goal of pca is to provide a function for carrying out a principle components analysis and two plotting functions to visualize the results of the analysis.

## Installation

You can install the development version of pca from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("asmauger/bios625hw3")
```

For detailed information about how to use this package, please see the vignette and help pages. Below is a brief example that showcases the three function of this package: `pca`, `elbowplot`, and `scoreplot`.

## Example analyzing mtcars data with PCA

The `mtcars` dataset contains observations on 11 variables for 32 cars. We can use `pca` to find principal components and the variance explained by each component. 

```{r example pca}
library(pca)
data(mtcars)
cars_pca = pca(~mpg + disp + hp + drat + wt + qsec, data=mtcars)
summary(cars_pca)
```

An elbowplot (or screeplot) can help us visualize the information provided in the summary from `pca`.

```{r example elbowplot, fig.width=7}
elbowplot(cars_pca)
```

This plot shows us that the first 3 components capture most of the variation in the mtcars data.

A scoreplot allows us to visualize the mtcars data in just two dimensions using the first two principal components.

```{r example scoreplot, warning=FALSE, fig.width=8}
library(ggplot2)
library(patchwork)
plot1 = scoreplot(cars_pca, grouping=mtcars$am) + labs(col='transmission')
plot2 = scoreplot(cars_pca, grouping=mtcars$cyl) + labs(col='cylinders')
plot1 + plot2
```

After coloring by the transmission type (automatic or manual), we see that the second PC (y axis) separates the cars based on transmission type. The plot coloring by number of cylinders shows us that the first component separates the cars by number of cylinders.
